Inferring probabilistic stellar rotation periods using Gaussian processes

### 2017-02-17 documents/ms.tex
removing comments
adding words to the conclusion

\newcommand{\lnacfMAD}{0.04}
\newcommand{\acfMAD}{0.60}
\newcommand{\percentacfMAD}{3.65}
\newcommand{\lnpgramMAD}{0.04}
\newcommand{\pgramMAD}{0.98}
\newcommand{\percentpgramMAD}{4.32}
\newcommand{\lngpMADnp}{0.02}
\newcommand{\gpMADnp}{0.48}
\newcommand{\percentgpMADnp}{2.45}
\newcommand{\lngpMAD}{0.02}
\newcommand{\gpMAD}{0.39}
\newcommand{\percentgpMAD}{1.96}

\newcommand{\lntelavivMAD}{0.02}
\newcommand{\telavivMAD}{0.34}
\newcommand{\percenttelavivMAD}{1.75}

\newcommand{\acfRMS}{1.75}
\newcommand{\telavivRMS}{25.09}
\newcommand{\pgramRMS}{19.50}
\newcommand{\gpRMSnp}{7.48}
\newcommand{\gpRMS}{7.12}

@@

We have attempted to recover the rotation periods of \naigrain\ simulated
\kepler-like light curves for solid-body rotators \citep{Aigrain2015} using
three different methods: a new Gaussian process method,
an authocorrelation function method, and a Lomb-Scargle periodogram method.
We demonstrate that the GP method produces the most accurate rotation periods
of the three techniques, providing a large improvement over the LS periodogram
method and a moderate one over our basic implementation of the ACF method.
We also find that our basic ACF method which is similar to methods implemented
in the literature, often slightly under-predicts the rotation period due to a
subtle feature of its peak detection algorithm.
Comparing our results with the results of the \citet{Aigrain2015} Tel Aviv
team who implement the method of \citet{Mcquillan2013}, a slightly different
version of the ACF method to the one we use, we find that their results are
slightly more precise than the GP method results overall.
They do however have a larger number of significant outliers.
We measured the rotation periods of \nkoimcq\ \kepler\ objects of interest
using the GP method, and find that these results compare well to those
measured previously by \citet{Mcquillan2013}.
The good agreement between these two sets of results demonstrates that the GP
method works well on real \kepler\ data.

Unlike the ACF and L-S periodogram methods, the GP method provides posterior
PDF samples which can be used to estimate rotation periods uncertainties.
Although an improvement on competing methods, these uncertainties are still
underestimated in many cases because the GP model is an approximate, effective
model. \tdmcomment{Can we make a statement something like ``we believe the method underestimates
uncertainties by about a factor of three'' or something like that?  This
matches with 66\% of the fits being within ``3$\sigma$''.
Might be good to say something like this, both in the discussion and here.}
% in many cases the posterior PDFs of the parameters are
% multi-modal.
% If it were possible to run an MCMC sampler for an infinite number of steps,
% the entire posterior PDF would eventually be sampled, and a more accurate
% uncertainty would result.
% Unfortunately, due to realistic computational resources and the limitations of
% MCMC sampling techniques for multi-modal posteriors, we are not always able to
% map out the entire posterior PDFs.
% \tdmcomment{I don't think this is the problem;
% I think the fits that we judge to have converged have actually converged.}
In addition, although the GP model used here is clearly a good one, it is
still only an effective model, not an accurate physical model.
It can only capture the posterior PDF of the periodic component of the
covariance matrix, not the actual rotation period of a physical star.

The main aim of this work was to develop a probabilistic rotation period
inference method.
Probabilistic periods are necessary for hierarchical Bayesian inference,
particularly when performing population analysis.
This new GP method is capable of generating a catalogue of probabilistic
rotation periods which could, for example, reveal the period distribution
(and therefore potentially the age distribution) of stars in the Milky Way.
There is still room for improvement since the uncertainty estimates are not
yet truly representative and it is still only an `effective' model, not a
physical one.
However, it is probabilistic and provides more accurate periods than
alternative methods.
We therefore argue that the GP method is the best method for rotation period
inference currently available.
The code used in this project is available at
\url{https://github.com/RuthAngus/GProtation/}.

% acknowledgements
This research was funded by the Simons Foundation and the Leverhulme Trust.
TDM is supported by NASA grant NNX14AE11G, and acknowledges the
hospitality of the Institute for Advanced Study,
where part of this work was completed.
VR thanks Merton College and the National Research Foundation of South Africa
for financial support.
Some of the data presented in this paper were obtained from the Mikulski
Archive for Space Telescopes (MAST).
STScI is operated by the Association of Universities for Research in
Astronomy, Inc., under NASA contract NAS5-26555.
Support for MAST for non-HST data is provided by the NASA Office of Space
Science via grant NNX09AF08G and by other grants and contracts.
This paper includes data collected by the Kepler mission. Funding for the
Kepler mission is provided by the NASA Science Mission directorate.

% \appendix

% \section{An ACF-informed Prior for Rotation Period}

% The autocorrelation function (ACF) has proven to be very useful
% for measuring the rotation periods of stars from time-series
% photometry.
% However, as discussed in the body of this paper, the
% ACF-based method has several shortcomings,
% most notably the inability to deliver uncertainties, but also
% the necessity of several heuristic choices,
% such as a timescale on which to smooth the ACF,
% how to define a peak, whether the first or second peak
% gets selected, and what constitutes a secure detection.
% While this paper presents a new MCMC-based method
% that avoids these shortcomings,
% it seems prudent to still use information available from the ACF.
% This Appendix describes how we use the ACF to define a prior on period.

% Our strategy here is not to attempt to decide which single
% peak in the ACF best represents the true rotation period,
% but rather to identify several \emph{candidate} periods and define
% a weighting scheme in order to create a non-commital, though useful,
% multimodal prior.  While this procedure does not avoid heuristic choices,
% the fact that its goal is simply to create a \emph{prior} for a
% probabilitistically justified inference scheme rather than
% to identify a single correct period softens the potential impact
% of these choices.

% As another innovation beyond what ACF methods in the literature have
% presented, we also bandpass filter the light curves (using a 5th order
% Butterworth filter, as implemented in \texttt{SciPy}) before calculating the
% autocorrelation.  This suppresses power on timescales shorter than a chosen
% minimum period $P_{\rm min}$ and longer than a chosen maximum $P_{\rm max}$,
% producing a cleaner autocorrelation signal than an unfiltered light curve.

% We use the following procedure to construct a prior for rotation period
% given a light curve:
% \begin{enumerate}
% \item{For each value of $P_i$, where $i = \{1, 3, 5, 10, 30, 50, 100\}$\,d,
% we apply a bandpass filter to the light curve using $P_{\rm min}=0.1$\,d
% and $P_{\rm max} = P_i$.  We then calculate the ACF of the filtered
% light curve out to a maximum lag of $2P_i$ and smooth it with a boxcar
% filter of width $P_i/10$.}

% \item{We identify the time lag corresponding to the
% first peak of each of these ACFs, as well as the first peak's
% trough-to-peak height, creating a set of candidate periods
% $T_i$ and heights $h_i$.}

% \item{We assign a quality metric $Q_i$ to each of these candidate
% periods, as follows.  First, we model the ACF as a
% damped oscillator with fixed period $T_i$:
% \begin{equation}
% y = A e^{-t/\tau} \cos{\frac{2\pi t}{T_i} },
% \end{equation}
% where and $t$ is the lag time,
% and find the best-fitting parameters $A_i$ and $\tau_i$ by a non-linear
% least squares minimization procedure.  We then define the
% following heuristic quality metric:
% \begin{equation}
% \label{eq:quality}
% Q_i = \left(\frac{\tau_i}{T_i}\right) \left(\frac{N_i h_i}{R_i}\right),
% \end{equation}
% where $h_i$ is the height of the ACF peak at $T_i$,
% $N_i$ is the length of the lag vector in the ACF (directly proportional
% to the maximum allowed period $P_i$),
% and $R_i$ is the sum of squared residuals between the
% damped oscillator model and the actual ACF data.  The idea behind this
% quality metric to give a candidate period a higher score if

%     \begin{enumerate}[(a)]

%     \item{it has many regular sinusoidal peaks, such that the decay
%         time $\tau_i$ is long compared to the oscillation period $T_i$,}
%     \item{the ACF peak height is high, and}
%     \item{the damped oscillator model is a good fit (in a $\chi^2$
%       sense) to the ACF, with extra bonus for being
%       a good fit over more points (larger $N_i$, or longer $P_i$).}

%     \end{enumerate}
% }

% \item{Given this set of candidate periods $T_i$ and quality metrics $Q_i$,
% we finally construct a multimodal prior on the $P$ parameter of the GP
% model as a weighted mixture of Gaussians:
% \begin{equation}
% \label{eq:mixture}
% p(\ln P) = \frac {\displaystyle \sum_i Q_i \left(0.9\mathcal N(\ln T_i, 0.2) +
%                                           0.05\mathcal N(\ln (T_i/2), 0.2) +
%                                           0.05\mathcal N(\ln (2 T_i), 0.2) \right)}
%                 {\sum_i Q_i}.
% \end{equation}
% That is, in addition to taking the candidate periods themselves as mixture
% components, we also mix in twice and half each candidate period at a lower level,
% which compensates for the possibility that the first peak in the ACF may actually
% represent half or twice the actual rotation period.  The period width of 0.2 in
% log space (corresponding to roughly 20\% uncertainty) is again a heuristic choice,
% balancing a healthy specificity with the desire to not have the results of
% the inference overly determined by the ACF prior.
% }

% Incidentally, while we use the procedure described here to create a prior on
% $P$ which we use while inferring the parameters of the quasi-periodic GP model,
% this same procedure may also be used in the service of a
% rotation-period estimating procedure all on its own, perhaps being even more
% robust and accurate than the traditional ACF method.  We leave exploration of
% this possibility to future work.

>>>>

\newcommand{\lnacfMAD}{0.14}
\newcommand{\acfMAD}{2.36}
\newcommand{\percentacfMAD}{13.02}
\newcommand{\acfRMS}{26.09}

\newcommand{\lnpgramMAD}{0.05}
\newcommand{\pgramMAD}{1.13}
\newcommand{\percentpgramMAD}{4.70}
\newcommand{\pgramRMS}{18.98}

\newcommand{\lngpMADnp}{0.02}
\newcommand{\gpMADnp}{0.48}
\newcommand{\percentgpMADnp}{2.45}
\newcommand{\gpRMSnp}{7.48}

\newcommand{\lngpMAD}{0.02}
\newcommand{\gpMAD}{0.39}
\newcommand{\percentgpMAD}{1.96}
\newcommand{\gpRMS}{7.12}

\newcommand{\lntelavivMAD}{0.02}
\newcommand{\telavivMAD}{0.34}
\newcommand{\percenttelavivMAD}{1.75}
\newcommand{\telavivRMS}{25.09}

@@

We have attempted to recover the rotation periods of \naigrain\ simulated
\kepler-like light curves for solid-body rotators \citep{Aigrain2015} using
three different methods: a new Gaussian process method,
an authocorrelation function method, and a Lomb-Scargle periodogram method.
We demonstrate that the GP method produces the most accurate rotation periods
of the three techniques, providing a large improvement over the LS periodogram
method and a moderate one over the ACF method.
We also find that the ACF method often slightly under-predicts the rotation
period due to a feature of its peak detection algorithm.
Comparing our results with the results of the \citet{Aigrain2015} Tel Aviv
team who implement the method of \citet{Mcquillan2013}, a slightly different
version of the ACF method to the one we use, we find that their results are
slightly more precise than the GP method results overall.
They do however have a larger number of significant outliers.
We measured the rotation periods of \nkoimcq\ \kepler\ objects of interest
using the GP method, and find that these results compare well to those
measured previously by \citet{Mcquillan2013}.
The good agreement between these two sets of results demonstrates that the GP
method works well on real \kepler\ data.

Unlike the ACF and LS periodogram methods, the GP method provides posterior
PDF samples which can be used to estimate rotation periods uncertainties.
In general it provides more accurate rotation periods than either method and
produces significantly fewer outliers.
The GP method can also be applied to non-uniformly sampled light curves,
unlike the ACF method.

Although an improvement on competing methods, the uncertainties are still
underestimated by the GP method in many cases because it is an approximate
model.
Although the GP model used here is clearly a good one, as it has been
demonstrated to reliably recover rotation periods, it is still only an
effective model, not an accurate physical model.
It captures the posterior PDF of the periodic component of the covariance
matrix of a time series, not the actual rotation period of a physical star.
The marginalised posterior PDF does not therefore exactly reflect the
probability of a rotation period, given the data.
Similarly, the uncertainty, calculated from the 16th and 84th percentiles of
the posterior PDF, is not exactly the standard deviation of a physical
rotation period.

Although underestimated uncertainties are a drawback of this method, they are
at least well motivated and well defined which is not the case for
uncertainties calculated from the ACF and LS periodogram methods.
The main aim of this work was to develop a probabilistic rotation period
inference method.
Probabilistic periods are necessary for hierarchical Bayesian inference,
particularly when performing population analysis.
This new GP method is capable of generating a catalogue of probabilistic
rotation periods which could, for example, reveal the period distribution
(and therefore potentially the age distribution) of stars in the Milky Way.
There is still room for improvement since the uncertainty estimates are not
yet truly representative and it is still only an `effective' model, not a
physical one.
However, it is probabilistic and provides more accurate periods than
alternative methods.
We therefore argue that the GP method is the best method for rotation period
inference currently available.
The code used in this project is available at
\url{https://github.com/RuthAngus/GProtation/}.

% acknowledgements
This research was funded by the Simons Foundation and the Leverhulme Trust.
TDM is supported by NASA grant NNX14AE11G, and acknowledges the
hospitality of the Institute for Advanced Study,
where part of this work was completed.
VR thanks Merton College and the National Research Foundation of South Africa
for financial support.
Some of the data presented in this paper were obtained from the Mikulski
Archive for Space Telescopes (MAST).
STScI is operated by the Association of Universities for Research in
Astronomy, Inc., under NASA contract NAS5-26555.
Support for MAST for non-HST data is provided by the NASA Office of Space
Science via grant NNX09AF08G and by other grants and contracts.
This paper includes data collected by the Kepler mission. Funding for the
\Kepler\ mission is provided by the NASA Science Mission directorate.
